---
title: "ESM 244 Lab 2"
author: "Allison Horst"
date: "December 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Lab 2 Objectives

- Dealing with function conflicts (function name in multiple packages)
- Ordinal logistic regression
- Principle components analysis example + biplot
- Shiny example

###0. Attach packages: 

Note: get devtools version of ggbiplot! 

    library(devtools)
    install_github("vqv/ggbiplot")

```{r, message = FALSE}

library(tidyverse)
library(janitor)
library(MASS)
library(RColorBrewer)
library(factoextra)
library(ggbiplot)
library(shinythemes)

```

Note: MASS package also has a 'select' function that will mask 'select' in dplyr - we don't want that! There are two things we can do: 

(1) State the package explicitly that the function comes from when using it, e.g. dplyr::select

(2) Generally override that the default select is from MASS as follows: 

    select <- dplyr::select
    
```{r, message = FALSE}
select <- dplyr::select
```

###1. OLR with political party (Open Science Framework: https://osf.io/8n6s4/)

####a. Get data (pol_party.csv) and do some wrangling

```{r, message = FALSE}

pol_party <- read_csv("pol_party.csv") %>% 
  clean_names() # janitor function for snake case

pol_df <- pol_party %>% 
  select(birth_year, gender, ethnicity, political_party) %>% 
  filter(birth_year > 1900, political_party != "NA", political_party <= 7) %>% 
  mutate(age = 2016 - birth_year)

# For political party: 1 = very liberal, 4 = moderate, 7 = very conservative

# Coerce to factors (including ordered factor for the ranking of political party)

pol_df$political_party <- factor(pol_df$political_party, ordered = TRUE, levels = c("1","2","3","4","5","6","7")) # Check class and levels to ensure

# Then update the gender and ethnicity variables to factors (unordered)

pol_df$ethnicity <- factor(pol_df$ethnicity)
pol_df$gender <- factor(pol_df$gender) 

# Then we can run ordinal logistic regression, but we should do some visualization first.

```

- Exploratory visualization

```{r, message = FALSE}

counts_eth <- pol_df %>% 
  group_by(ethnicity, political_party) %>% 
  tally()

counts_g <- pol_df %>% 
  group_by(gender, political_party) %>% 
  tally

# Exploratory graphs

ggplot(pol_df, aes(x = age, y = political_party)) +
  geom_point(alpha = 0.5, aes(color = ethnicity)) # no notable effect of age

ggplot(pol_df, aes(x = political_party)) +
  geom_histogram(aes(fill = ethnicity), stat = "count") +
  scale_fill_brewer(palette = "YlGnBu") +
  theme_light() +
  theme(legend.position = "NA") +
  facet_wrap(~ethnicity, scales = "free")

ggplot(pol_df, aes(x = political_party)) +
  geom_histogram(aes(fill = gender), stat = "count") +
  facet_wrap(~gender)

```

- Ordinal logistic regression using MASS::polr

```{r, message = FALSE}
# Ordinal logistic regression

pol_model <- polr(political_party ~ age + gender + ethnicity, data = pol_df) # Run OLR

summary(pol_model) # See model summary. Note: can convert this to p-values if you really MUST. 

exp(coef(pol_model)) # See exponentiated (odds) coefficients

```

- Probability predictions using the model 

```{r, message = FALSE}
# Find predictions for the existing row inputs
pred <- predict(pol_model, type = "probs")

# Bind together with original data rows
df <- data.frame(pol_df, pred)

# And then you could decide how to show this visually...

```

###2. World bank environment PCA (Principle components analysis)

- Data from the World Bank (<https://data.worldbank.org/topic/environment>)

a. Get the data and do some wrangling

```{r, message = FALSE}

countries <- read_csv("country_info.csv")
wb_df <- read_csv("wb_df.csv")

regions <- countries %>% 
  select(code, Region)

# Only use data for 2010 (make sure to check what each step does in wrangling)

wb_10 <- wb_df %>% 
  select(name, code, indicator, `2010`) %>% 
  filter(indicator == "CO2 emissions (kt)" |
           indicator == "Access to electricity (% of population)" |
           indicator == "Agricultural land (% of land area)" |
           indicator == "Methane emissions (kt of CO2 equivalent)" |
           indicator == "Total greenhouse gas emissions (kt of CO2 equivalent)" |
           indicator == "PM2.5 air pollution, mean annual exposure (micrograms per cubic meter)"
) %>% 
  dplyr::rename(value = `2010`) %>% 
  filter(value != "NA") %>% 
  full_join(regions) %>% 
  group_by(code) %>% 
  spread(indicator, value) %>% 
  select(-`<NA>`) %>% 
  ungroup() %>% 
  clean_names() %>% 
  filter(region == "North America" |
           region == "Latin America & Caribbean") %>% 
  dplyr::rename(electricity = access_to_electricity_percent_of_population, 
         agland = agricultural_land_percent_of_land_area,
         co2 = co2_emissions_kt,
         methane = methane_emissions_kt_of_co2_equivalent,
         ghg = total_greenhouse_gas_emissions_kt_of_co2_equivalent,
         air_poll = pm2_5_air_pollution_mean_annual_exposure_micrograms_per_cubic_meter) %>% 
  filter(electricity != "NA",
         agland != "NA",
         co2 != "NA",
         methane != "NA",
         ghg != "NA")

write.csv(wb_10, "wb_10.csv")

# Then just read in wb_10...but the wrangling code is there if you want to see it.

```

- Principal components analysis (stats::prcomp) with biplots

```{r, message = FALSE}

wb_pca <- prcomp(wb_10[4:8], scale = TRUE)
summary(wb_pca) # Explore the eigenvectors!

# ggbiplot for visualization
ggbiplot(wb_pca, labels = wb_10$name, groups = wb_10$region) +
  theme_bw()
  
# Alternatively, with factominer fviz functions: 
  
fviz_pca_biplot(wb_pca,
                col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
)

fviz_pca_ind(wb_pca,
             col.ind = "cos2", 
             gradient.cols = c("orange", "magenta", "blueviolet"),
             repel = TRUE
             )

```

More PCA in R examples: 

Clear R code: 
<https://www.datacamp.com/community/tutorials/pca-analysis-r>

Awesome visualization examples using factominer:
<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/>


###3. Shiny example: Marvel characters (create a new Shiny app)


